---
title: "Effects of Income on Average SAT score"
output: github_document
editor_options: 
  chunk_output_type: console
---

**This is an exploration of how average household income influences university entrance exam scores. Income data was sourced from the IPUMS NHGIS database while exam score data was sourced from the GOSA database (Georgia office of student achievement).**

```{r}
library(knitr)
library(tidyverse)
library(sf)
library(spatstat)
library(tmap)
library(car)

```
 
Import income data and sort into two different categories.
 
```{r}
income_data<-read.csv("data/complete analysis.csv")

#Household income < 25k/year
low_income<-income_data%>%
  select(School_DSTRCT_CD, SDUNI, SDUNIA, NAME_E,
         AJY9E001, AJY9E002, AJY9E003, AJY9E004, AJY9E005)%>%
  mutate(Count_less30k = AJY9E002 + AJY9E003 + AJY9E004 + AJY9E005)%>%
  mutate(pct_low_inc = (AJY9E002 + AJY9E003 + AJY9E004 + AJY9E005)/AJY9E001)
kable(head(low_income))

#Household income > 100k/year
high_income<-income_data%>%
  select(School_DSTRCT_CD, SDUNI, SDUNIA, NAME_E,
         AJY9E001, AJY9E014, AJY9E015, AJY9E016, AJY9E017)%>%
  mutate(Count_more100k = AJY9E014 + AJY9E015 + AJY9E016 + AJY9E017)%>%
  mutate(pct_high_inc = (AJY9E014 + AJY9E015 + AJY9E016 + AJY9E017)/AJY9E001)
kable(head(high_income))
```

Import SAT scores and filter so that only records for combined scores remain.

```{r}
raw_SAT_scores<-read.csv("data/All SAT scores GA 2011-2019.csv")%>%
  group_by(LONG_SCHOOL_YEAR, SCHOOL_DISTRCT_CD)%>%
  filter(TEST_CMPNT_TYP_CD == "Combined" | TEST_CMPNT_TYP_CD == "Combined Test Score")%>%
  filter(LONG_SCHOOL_YEAR == "2014-15" | LONG_SCHOOL_YEAR == "2015-16" | LONG_SCHOOL_YEAR == "2016-17" | LONG_SCHOOL_YEAR == "2017-18" | LONG_SCHOOL_YEAR == "2018-19")
kable(head(raw_SAT_scores))
  
new_scale_scores<-raw_SAT_scores%>%
  filter(LONG_SCHOOL_YEAR == "2015-16" | LONG_SCHOOL_YEAR == "2016-17" | LONG_SCHOOL_YEAR == "2017-18" | LONG_SCHOOL_YEAR == "2018-19")%>%
  group_by(LONG_SCHOOL_YEAR, SCHOOL_DISTRCT_CD, INSTN_NUMBER)%>%
  mutate(scale = "New Scale")
kable(head(new_scale_scores))

SAT_scores<-raw_SAT_scores%>%
  left_join(new_scale_scores)%>%
  replace_na(list(scale = "Old Scale"))
kable(head(SAT_scores))
```

Now convert scores on the old scale. The scale changed from 2400 to 1600 in the spring 2016. Create a conversion calculator.

```{r}
conversion_scale<-read.csv("data/Conversion chart new to old SAT dash separat.csv")
kable(head(conversion_scale))

new_score_calc<-function(old_score){
  max_range<-conversion_scale %>%
    filter(Old_Max>old_score)%>%
    filter(Old_Max == min(Old_Max))
data.frame(score=old_score,new_score=max_range[,1])
}
test<-new_score_calc(1374)

#Conversion of institutional scores
inst_scores<-SAT_scores%>%
  filter(LONG_SCHOOL_YEAR == "2014-15")%>%
  mutate(score = as.numeric(INSTN_AVG_SCORE_VAL))%>%
  filter(score > 0)

score_transform<-map_df(inst_scores$score, new_score_calc)

scores<-as.numeric(score_transform$INSTN_AVG_SCORE_VAL)

final_converted_scores<-inst_scores%>%
  left_join(score_transform)

all_old_scores_final_instn<-final_converted_scores%>%
  full_join(new_scale_scores)
kable(head(all_old_scores_final_instn))

#Conversion of district scores
DSTRCT_scores<-SAT_scores%>%
  filter(LONG_SCHOOL_YEAR == "2014-15")%>%
  mutate(score = as.numeric(DSTRCT_AVG_SCORE_VAL))%>%
  filter(score > 0)

score_transform2<-map_df(DSTRCT_scores$score, new_score_calc)

scores2<-as.numeric(score_transform2$DSTRCT_AVG_SCORE_VAL)

final_converted_scores2<-DSTRCT_scores%>%
  left_join(score_transform2)

all_old_scores_final_DSTRCT<-final_converted_scores2%>%
  full_join(new_scale_scores)
kable(head(all_old_scores_final_DSTRCT))
```

Get the average of all years SAT scores for each school district. Join the income data to the SAT scores according to SCHOOL_DISTRCT_CD

```{r}

#this data frame only contatains the score data from 2014-15. The score values for the rest of the years appear as NA
aggregate_scores_dist<-all_old_scores_final_DSTRCT%>%
  full_join(new_scale_scores)%>%
  group_by(LONG_SCHOOL_YEAR ,SCHOOL_DISTRCT_CD, DSTRCT_AVG_SCORE_VAL, scale, score, new_score)%>%
  summarise(District_avg_score = mean(new_score))%>%
  mutate(Aggregate_DSTRC_avg_Score = (mean(as.numeric(DSTRCT_AVG_SCORE_VAL) + new_score)/2))%>%
  ungroup()
  
kable(head(aggregate_scores_dist))
```

